---
title: "Assignment 1"
author: "Shreejoyee Nath"
date: "2026-01-19"
output: word_document
---

1. Report the “class” of the data set. How many rows and columns are in
this data set? What do the rows and columns represent?

```{r}
library(MASS)
data(Boston)
class(Boston)
dim(Boston)
str(Boston)

```
Class of the data set:
The data set is of class data.frame (pandas DataFrame equivalent).

Dimensions:
Number of rows: 506
Number of columns: 14

Interpretation:
Rows represent 506 suburbs of Boston.
Columns represent different socio-economic, environmental, and housing-related variables, such as crime rate, tax rate, pupil–teacher ratio, pollution levels, and median value of owner-occupied homes.

2. Create a smaller data set with the variables median value of owner-occupied
homes, per capita crime rate, nitrogen oxides concentration, proportion of
blacks and percentage of lower status of the population. Choosing median
value of owner occupied homes as the response and the rest as the pre-dictors, make scatter plots of the response versus each predictor. Present the scatter plots in different panels of the same graph. Comment on your findings.

```{r}
Boston_sub <- Boston[, c("medv", "crim", "nox", "black", "lstat")]
par(mfrow = c(2, 2))

plot(Boston_sub$crim, Boston_sub$medv,
     xlab = "Crime rate",
     ylab = "Median house value",
     main = "medv vs crim")

plot(Boston_sub$nox, Boston_sub$medv,
     xlab = "Nitrogen oxides",
     ylab = "Median house value",
     main = "medv vs nox")

plot(Boston_sub$black, Boston_sub$medv,
     xlab = "Proportion of blacks",
     ylab = "Median house value",
     main = "medv vs black")

plot(Boston_sub$lstat, Boston_sub$medv,
     xlab = "Lower status population",
     ylab = "Median house value",
     main = "medv vs lstat")

```


3. Which suburb of Boston has lowest median value of owner-occupied homes?
What are the values of the other predictors mentioned in (2), for that suburb. How do these values compare to the overall ranges for those pre-dictors? Comment on your findings. Hint: Mention which percentile these values belong to.

```{r}
min_medv <- Boston[which.min(Boston$medv), ]
min_medv
apply(Boston[, c("crim","nox","black","lstat")], 2,
      function(x) ecdf(x)(min_medv[colnames(x)]) * 100)

```
The percentage of lower-status residents and the crime rate are both above the 95th percentile in the suburb with the lowest median value of owner-occupied residences. This explains why the value of the house is so low.

4. Does any suburb of Boston stand out for having notably high crime rates,
tax rates, or pupil–teacher ratios? Hint: Use a boxplot to detect any
outliers. If so, identify the suburbs that show the outlier values.

```{r}
par(mfrow = c(1, 3))

boxplot(Boston$crim, main = "Crime rate", ylab = "crim")
boxplot(Boston$tax, main = "Tax rate", ylab = "tax")
boxplot(Boston$ptratio, main = "Pupil-Teacher Ratio", ylab = "ptratio")
boxplot.stats(Boston$crim)$out
boxplot.stats(Boston$tax)$out
boxplot.stats(Boston$ptratio)$out

```
The boxplots show a number of extreme anomalies in the tax and crime rates. While the student-teacher ratio displays fewer and less extreme outliers, a few suburbs have abnormally high crime and tax values.
