---
title: "Predictive Assignment 3"
author: "Shreejoyee Nath_747"
date: "2026-02-19"
output: html_document
---

5. Problem to demonstrate the utility of nonlinear regression over linear regression

Get the fgl data set from “MASS” library.

```{r Vehicle window glass}
library(MASS)
data("fgl")
View(fgl)
```

(a) Considering the refractive index (RI) of “Vehicle Window glass” as the variable of interest and assuming linearity of regression, run multiple linear regression of RI on different metallic oxides. From the p value, report which metallic oxide best explains the refractive index.

```{r linear model}
fgl_vwg = subset(fgl,fgl$type == "Veh")
model_RI_vwg = lm(RI~Na+Mg+Al+Si+K+Ca+Ba+Fe, data = fgl_vwg)
summary(model_RI_vwg)

```
Interpretation:
Oxides iron are best,for explaining the RI of Vehicle window glass. 

(b) Run a simple linear regression of RI on the best predictor chosen in (a).

```{r lin model on best predictors}
model_fgl_best_oxides = lm(RI~Fe,data=fgl_vwg)
summary(model_fgl_best_oxides)
```
Interpretation:

$R^2$ = 0.2097

It explains 21% of the variance in RI of VWG.

(c) Can you further improve the regression of the refractive index of “Vehicle
Window glass” on the predictor chosen by you in part (a)? Give the new fitted
model and compare its performance with the model in (b).

```{r better model}
model_better = lm(RI~Fe+I(Fe**2),data = fgl_vwg)
summary(model_better)
```
Interpretation:

$R^2$ = 0.3554. 

it explains 35% variance of RI in VWG.

